---
title: "Statistical inference"
author: "Moisès Bernabeu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  # pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = '100%', fig.width = 5,
                      fig.height = 4, out.width='70%', fig.align = 'center')

library(kableExtra)
library(latex2exp)
```

## Inference concept

### Coin toss heads and tails probability inference

**Definition**: We define a random variable $X$ as the function that maps every element of the sample space $\Omega$ to a measurable space $E$. It is  $X:\Omega\rightarrow E$.

**Example**: When we have discrete data, like a factor or number of rains per year, $E$ are natural numbers, then $E = \mathbb{N}$ and $X : \Omega \rightarrow \mathbb{N}$. When the variable is continuous, like the height of women, $E = \mathbb{R}^{+}$, then $X : \Omega \rightarrow \mathbb{R}^{+}$, finally, if the variable can take any real values $E = \mathbb{R}$, then $X : \Omega \rightarrow \mathbb{R}$.

Let $\mathcal{D} = (d_1, d_2, \dots, d_n)$ be the **random variable** of the event: having a head on a coin toss trial, where $n$ is the sample size and each $d_i$ will be an element of the random variable. If sample space is $\Omega = \{T, H\}$, and the success of this trial is "having a head" (like we defined the random variable). $d_i = 1$ in $\mathcal{D}$ will be denoting having a head, and $d_i = 0$ a tile. Then, from an experiment of $n = 10$, we should get a data vector like:
\[
\mathcal{D} = (1, 0, 0, 0, 1, 0, 0, 1, 1, 0).
\]

The distribution of this kind of data (success and failure) is called Bernoulli. It is a particular case of Binomial distributed data. Binomial distribution explains the number of successes in an $n$ length sequence of trials (e.g. number of baskets scored in 10 shots). When the number of trials is 1, the distribution is called Bernoulli, then it is the success or failure of a single shot, coin toss, etc.

## Variables
There exist two types of quantitative variables:

1. **Discrete variables**: are the variables that takes integer values, $\mathbb{Z}$. For instance, the number of heads in 10 coin toss trials, the variable ranges 0 to 10, but 1.5 is an impossible outcome.
2. **Continuous variables**: are the expressed in real numbers variables, $\mathbb{R}$. The diameter of a pine tree, for instance, it ranges positive reals, and 20.34 is a possible value of the variable.

Random variables, have an associated uncertainty, the mathematical framework for this uncertainty is provided by the distribution that the variable follows. We define distribution as the mathematical formulas and structures that associate each possible event or a set of events to their probabilities.

### Distribution characterisation
#### Discrete variables
Distributions are characterized by two functions, in discrete variables, the **Probability Mass Function** (PMF, or simply the probability function) associates a probability to an event of the sample using the random variable. The probability function is denoted $f(x)$ and its definition is $f(x_i) = P(X = x_i)$, where $x_i$ is an element of the random variable $X$ that is linked to an element of the sample space. The other function that defines the distribution is the **Cumulative Distribution Function** (CDF, or distribution function), as it name says, it accumulates the values of $P(X = x_i)$, then
\[
F(x) = P(X \leq x_i) = \sum_{i = 0}^{n} P(X = x_i).
\]

```{r, echo=FALSE, fig.width=9, fig.height=4, out.width='80%'}
par(mfrow = c(1, 2))

pmf <- dbinom(0:10, 10, 0.6)
names(pmf) <- 0:10

barplot(pmf, xlab = 'x', ylab = TeX('$f(X = x) = P(X = x)'),
        main = 'Probability Mass Function')

cdf <- pbinom(0:10, 10, 0.6)
names(cdf) <- 0:10

plot(cdf, pch = 20, xlab = 'x', ylab = TeX('$F(X = x) = P(X \\leq x)$'),
     main = 'Cumulative Distribution Function')
for (i in 1:length(cdf)) {
  segments(x0 = i, x1 = i + 1, y0 = cdf[i], cdf[i])
}
```

In above figure it is seen the distribution function saturates at 1, it is because the probability ranges 0 to 1, and the probability is maintained until the next event of the sample space.

#### Continuous variables
In continuous variables there exist the **Probability Density Function** (PDF, or density function). It does not give us the probability at each point (because its probability is 0 since in a continuous variable there exist infinite set of possible values, then the probability of a point is $1 / \infty = 0$), but the relative likelihood that a test variable value is close to this distribution, in this function, the probabilities are areas under the PMF curve, we can talk about probabilities that the value is between 2 values $P(x_i < X < x_j)$, is greater $P(X < x_i)$ or lower $P(X < x_i)$ (note we never use the equal, since it is 0, it would be caught, but it will not change the probability). As discrete variables, it is also characterized by the **Cumulative Distribution Function** (CDF, or distribution function), in the continuous case
\[
F(X = x_i) = P(X < x_i) = \int_{-\infty}^{x_i} f(X = x) \, dx
\]

```{r, echo=FALSE, fig.width=9, fig.height=4, out.width='80%'}
par(mfrow = c(1, 2))

pdff <- dnorm(-400:400/100)

plot(-400:400/100, pdff, type = 'l', xlab = 'x', ylab = TeX('$f(X = x)$'),
     main = 'Probability Density Function')

cdf <- pnorm(-400:400/100)

plot(cdf, type = 'l', xlab = 'x', ylab = TeX('$F(X = x) = P(X < x)$'),
     main = 'Cumulative Distribution Function')
```

## Bernoulli distribution
As described before, the coin toss experience is a Bernoulli, which explains the probability of success and failure. We define a Bernoulli process with the Probability Mass Function
\[
P(\mathcal{D} = d_i; p) = p^{d_i}(1 - p)^{1 - d_i}.
\]
Note when $d_i = 1$ (it is a head), the term $p^{d_i}(1 - p)^{1 - d_i} = p$, the probability of having a head, and, analogously, when $d_{i} = 0$, $p^{d_i}(1 - p)^{1 - d_i} = 1 - p = q$, which is the probability of having a tail. If we are searching for a whole population, we want to estimate the $\pi$ parameter, which is the $p$ but for the population.

## Non-parametric estimate
The number of successes divided by the number of events, will give a non parametric estimate of the $p$, this means that data is not modelled using distributions having parameters.

```{r}
# Coin toss with non-parametric calculation ----
# Setting real parameter
rpi <- 0.3

npest <- c()
for (i in c(5, 10, 50, 100, 500, 1000)) {
  # Setting a seed to have always the same simulation results
  set.seed(0382098)
  
  # Getting a Bernoulli (binomial with sample size = 1) sample of i elements
  # and pi = real pi. 1 = head and 0 tail
  d <- rbinom(i, 1, rpi)
  
  npest <- c(npest, sum(d) / length(d))
}

npest
```

## Maximum Likelihood Inference
Likelihood methods are based on the maximum likelihood function, we denote it $L$, and it means the probability of having given data when the parameter gets a value, then
\[
L(\pi) = P(\mathcal{D} \, \vert \, \pi) = \prod_{i = 1}^{n} P(\mathcal{D} = d_i \, \vert \, \pi) = \prod_{i = 1}^{n} \pi^{d_i}(1 - \pi)^{1 - d_i}
\]
is the likelihood for $\mathcal{D}$ when the parameter of the Bernoulli distribution is $\pi$. If we do this for all possible values of the $\pi$ parameter (i.e. $[0,1]$), we will have the likelihood function for all the possible parameters, we define, then, the Maximum Likelihood Estimate (MLE) as the parameter value which maximises the likelihood value. From this MLE, we obtain a distribution parameter estimator, which usually is denoted $\hat{\pi}$.

Next code and figure show the experiment likelihood computations. We started from a given real parameter (which we want to infer, `rpi`), and we simulate a data set of size $n$, we calculate the likelihood for each possible $\pi \in [0, 1]$. We plotted the function and its MLE as vertical line and its confidence interval. $\pi$ is assumed to be asymptotically normal for large $n$s, then, its confidence interval at $\alpha$ level will be given by
\[
\hat{\pi} \pm z_{1 - \alpha} \sqrt{\frac{\hat{\pi}(1-\hat{\pi})}{n}},
\]
this works better for samples higher with $n \geq 30$.

```{r, fig.width=11.5, fig.height=6.33, out.width='100%'}
# Coin toss with ML ----
# Setting real parameter
rpi <- 0.3

par(mfrow = c(2, 3))
for (i in c(5, 10, 50, 100, 500, 1000)) {
  # Setting a seed to have always the same simulation results
  set.seed(0382098)
  
  # Getting a Bernoulli (binomial with sample size = 1) sample of i elements
  # and rpi = real pi. 1 = head and 0 = tail
  d <- rbinom(i, 1, rpi)
  
  # Generating the estimate pi values and computing likelihood as the product
  # of the probability mass function for our data (d sample) evaluated at pi
  # parameter
  pi <- 0:1000/1000
  like <- c()
  for (i in 1:length(pi)) {
    # Computing likelihood
    like <- c(like, prod(dbinom(d, 1, pi[i])))  # 1 - pi[i] will plot q MLE
  }
  
  # Getting the Maximum Likelihood estimate
  MLE <- pi[which.max(like)]
  
  # Getting confidence interval
  epi <- sum(d) / length(d)
  CI95 <- epi + 1 * qnorm(c(0.025, 0.975)) * sqrt(epi * (1 - epi) / length(d))
  
  # Plotting the likelihood function for the paramiter with d data and its MLE
  plot(pi, like, type = 'l', main = paste('n =', length(d)),
       xlab = TeX('$\\pi = P(D = 1)$'), ylab = 'Likelihood')
  abline(v = MLE, lty = 4, col = 'steelblue')
  abline(v = CI95, lty = 4, col = 'darkorange3')
  legend('topright', inset = 0.02,
         legend = c(paste('MLE =', round(MLE, 2)), 'CI 95 %'),
         col = c('steelblue', 'darkorange'), lty = 4)
}
```

While the sample size grows, the likelihood uncertainty on the parameter space is reduced (dispersion of the likelihood for greater $n$ samples is lower) and it confidence interval is lower. This means that higher amount of data will give less uncertain parameter. Remember that the confidence interval is not a probability, but the number of times a parameter would be in the interval if we repeat the experiment 100 times.

## Bayesian Inference
The main goal of Bayesian inference is the same than frequentist inference (ML), to learn about a population from a subset of it (the sample). The difference lies in the conception of the probability, not at mathematical level, but how to interpret it. In fraquentist statistics, the probability is undesrtood as the limit of the relative frequency ($P(A) = \lim_{n \rightarrow m} f(A) / n$, where $m$ is the population size). Nevertheless, the Bayesian concept of probability is linked to the uncertainty of an event.

Bayes' theorem provides the possibility to calculate conditional probabilities. Let $A$ and $B$ be random events, the probability of $A$ given (having occurred) $B$ is
\[
P(A \, \vert \, B) = \frac{P(B \, \vert \, A) P(A)}{P(B)} \propto P(B \, \vert \, A) P(A).
\]

Using this theorem we can structure a learning from data process. It is the Bayesian inference process, and it has these components:

1. **Quantity of interest**: the object of our interest, in this case the parameter $\pi$, which is the probability of having a head in a coin toss.
2. **Prior distribution**: previous known information about the quantity of interest, we could say we actually know the probability of a head or a tail is both 0.5, this allows us to design a prior distribution with more density in 0.5. $P(\pi)$
3. **Prior predictive distribution**: the prior probability that the quantity of interest is greater, lower or is inside an interval before doing the experiment, only with the prior distribution.
4. **Likelihood function**: merges the data with our interest quantity. $L(\pi) = P(\mathcal{D} \, \vert \, \pi)$.
5. **Posterior distribution**: Baye's rule updated probability for the interest quantity. This update is made from the data through the likelihood function, it is denoted by $P(\pi \, \vert \, \mathcal{D}) \propto P(\mathcal{D} \, \vert \, \pi) P(\pi)$.
6. **Posterior predictive distribution**: the posterior probability of having a value greater, lower or inside an interval for the interest quantity.

```{r, fig.width=11.5, fig.height=6.33, out.width='100%'}
alpha <- c(0.5, 0.5, 1, 1, 2, 5)
beta <- c(0.5, 1, 1, 1.5, 2, 2)

par(mfrow = c(2, 3))
for (j in 1:length(alpha)) {
  i <- 10
  # Setting a seed to have always the same simulation results
  set.seed(0382098)
  
  # Getting a Bernoulli (binomial with sample size = 1) sample of i elements
  # and rpi = real pi. 1 = head and 0 = tail
  d <- rbinom(i, 1, rpi)
  
  # Generating the estimate pi values and computing likelihood as the product
  # of the probability mass function for our data (d sample) evaluated at pi
  # parameter
  pi <- 0:1000/1000
  
  # Generating a prior which will be used to the Bayesian inference
  prior <- dbeta(seq(0, 1, length.out = length(pi)), alpha[j], beta[j])
  
  like <- c()
  for (i in 1:length(pi)) {
    # Computing likelihood
    like <- c(like, prod(dbinom(d, 1, pi[i])))
  }
  
  # Removing infinity values (they are because of the existence of asymptotes
  # in 0 or 1)
  infprior <- which(prior == Inf)
  if (length(infprior) > 0) {
    prior <- prior[-infprior]
    like <- like[-infprior]
    pi <- pi[-infprior]
  }
  
  # Computing weighted likelihood
  whlike <- prior * like
  
  # Computing normalising bayesian constant by the total probability theorem
  nmctnt <- sum(prior * like)
  
  post <- whlike / nmctnt
  
  # Plotting the likelihood function for the paramiter with d data and its MLE
  if (mean(prior) != 1) {
    plr <- range(c(like, post, prior))
    plot(pi, prior, type = 'l', col = 'darkorange3', lty = 2, lwd = 1.5,
         ylab = 'Relative density', xlab = TeX('$\\pi$'),
         yaxt = 'n', ylim = plr)
    lines(pi, like * max(plr) / max(like), col = 'steelblue', lty = 4, lwd = 1.5)
    lines(pi, post * max(plr) / max(post), type = 'l', col = 'brown3', lwd = 2)
    legend('topright', inset = 0.02, lty = c(2, 4, 1), lwd = 2,
           col = c('darkorange3', 'steelblue', 'brown3'),
           legend = c('Prior', 'Likelihood', 'Posterior'))
    title(paste0('Be(', alpha[j], ', ', beta[j], '), n = 10'))
  } else {
    plr <- c(0, 2)
    plot(pi, prior, type = 'l', col = 'darkorange3', lty = 2, lwd = 1.5,
         ylab = 'Relative density', xlab = TeX('$\\pi$'),
         yaxt = 'n', ylim = plr)
    lines(pi, like * 2 / max(like), col = 'steelblue', lty = 4, lwd = 1.5)
    lines(pi, post * 2 / max(post), type = 'l', col = 'brown3', lwd = 2)
    legend('topright', inset = 0.02, lty = c(2, 4, 1), lwd = 2,
           col = c('darkorange3', 'steelblue', 'brown3'),
           legend = c('Prior', 'Likelihood', 'Posterior'))
    title(paste0('Be(', alpha[j], ', ', beta[j], '), n = 10'))
  }
}
```

```{r, fig.width=11.5, fig.height=6.33, out.width='100%'}
par(mfrow = c(2, 3))
for (i in c(5, 10, 50, 100, 500, 1000)) {
  # Setting a seed to have always the same simulation results
  set.seed(0382098)
  
  # Getting a Bernoulli (binomial with sample size = 1) sample of i elements
  # and rpi = real pi. 1 = head and 0 = tail
  d <- rbinom(i, 1, rpi)
  
  # Generating the estimate pi values and computing likelihood as the product
  # of the probability mass function for our data (d sample) evaluated at pi
  # parameter
  pi <- 0:1000/1000
  
  # Generating a prior which will be used to the Bayesian inference
  prior <- dbeta(seq(0, 1, length.out = length(pi)), 2, 2)
  
  like <- c()
  for (i in 1:length(pi)) {
    # Computing likelihood
    like <- c(like, prod(dbinom(d, 1, pi[i])))
  }
  
  # Removing infinity values (they are because of the existence of asymptotes
  # in 0 or 1)
  infprior <- which(prior == Inf)
  if (length(infprior) > 0) {
    prior <- prior[-infprior]
    like <- like[-infprior]
    pi <- pi[-infprior]
  }
  
  # Computing weighted likelihood
  whlike <- prior * like
  
  # Computing normalising bayesian constant by the total probability theorem
  nmctnt <- sum(prior * like)
  
  post <- whlike / nmctnt
  
  # Plotting the likelihood function for the paramiter with d data and its MLE
  plr <- c(0, 2)
  plot(pi, prior, type = 'l', col = 'darkorange3', lty = 2, lwd = 1.5,
       ylab = 'Relative density', xlab = TeX('$\\pi$'),
       yaxt = 'n', ylim = plr)
  lines(pi, like * 2 / max(like), col = 'steelblue', lty = 4, lwd = 1.5)
  lines(pi, post * 2 / max(post), type = 'l', col = 'brown3', lwd = 2)
  legend('topright', inset = 0.02, lty = c(2, 4, 1), lwd = 2,
         col = c('darkorange3', 'steelblue', 'brown3'),
         legend = c('Prior', 'Likelihood', 'Posterior'))
  title(paste0('Be(2, 2), n = ', length(d)))
}
```